---
title: "Review analysis of mutations & peptides"
author: "Jacqueline Buros"
date: "6/16/2017"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warnings = FALSE)

library(readr)     # load data
library(survival)  # base survival (coxph, etc)
library(survminer) # better plotting of KM curves
library(tidyverse) # data manipulation

# load data
data_url <- 'https://raw.githubusercontent.com/hammerlab/paper-aocs-chemo-neoantigens/master/additional-files/Additional%20File%201.csv'
d <- readr::read_csv(data_url) 

## modified or mutated data stored in md
above_quantile <- function(x, q, ...) {
  x > quantile(x, probs = q, ...)
}

above_quantile_str <- function(x, q, ...) {
  threshold <- as.integer(quantile(x, probs = q, ...))
  ifelse(x > threshold, stringr::str_c('>', threshold), stringr::str_c('<=', threshold))
}

md <- d %>%
  dplyr::mutate(event = donor_vital_status == 'deceased',
                mutations_excl = ifelse(mutations >= 30000, NA, mutations)) %>%
  dplyr::filter(specific_treatment == 'primary/untreated' & tissue_type == 'solid') %>%
  tidyr::drop_na(event, donor_survival_time, tumour_stage, mutations) %>%
  dplyr::mutate_each(funs = funs(above_median = above_quantile(., q = 0.5, na.rm = T),
                                 above_median_str = above_quantile_str(., q = 0.5, na.rm = T),
                                 above_p75 = above_quantile(., q = 0.75, na.rm = T),
                                 above_p75_str = above_quantile_str(., q = 0.75, na.rm = T)
                                 ),
                     mutations, peptides, mutations_excl)

md_all <- d %>%
  dplyr::mutate(event = donor_vital_status == 'deceased') %>%
  tidyr::drop_na(event, donor_survival_time, tumour_stage, mutations) %>%
  dplyr::mutate_each(funs = funs(above_median = above_quantile(., q = 0.5),
                                 above_median_str = above_quantile_str(., q = 0.5),
                                 above_p75 = above_quantile(., q = 0.75),
                                 above_p75_str = above_quantile_str(., q = 0.75)
  ),
  mutations, peptides)

# mull model for plotting
m0 <- survival::survfit(Surv(time = donor_survival_time, event = event) ~ 1, data = md)

# null model for coxph
mph0 <- survival::coxph(Surv(time = donor_survival_time, event = event) ~ 1, data = md)
```


## Analysis of tumour stage

Plotting KM estimates by tumour stage

```{r plot-tumour-stage}
update(m0, . ~ . + tumour_stage, data = md) %>%
  survminer::ggsurvplot(., risk.table = TRUE, conf.int = TRUE) 
```

Fit coxph model 

```{r coxph-tumour-stage}
mph1 <- update(mph0, . ~ . + tumour_stage)
summary(mph1)
```

Test for non-proportional hazards over linear time

```{r zph-tumour-stage-linear}
cox.zph(mph1)
```

Test for non-proportional hazards over log-time

```{r zph-tumour-stage-log}
cox.zph(mph1, transform = log)
```

_ These both suggest that we don't have NPH in the unadjusted data, which means we will likely not need to stratify by tumour stage. Also suggests that some of the tumour stage association could be due to small sample sizes _

_ Still, the direction is pretty unexpected & given small sample size with stage IV I wonder if there isn't some selection bias for solid / untreated stage IV tumours._

Test for overly influential observations (on their effect on beta - so scale here is relative to beta estimate).

```{r influence-tumour-stage}
mph1.dfbeta <- residuals(mph1, type="dfbeta")
plot(mph1.dfbeta)
```


This caused me to wonder why some of these were so strongly negative -- is there heterogeneity in assessment of tumour stage?

Did some looking at tumour-grade & percentage cellularity, which didn't modify the influence of these observations. Also tried to look at larger (ascites / treated/relapse) data to see if there was selection bias, but tumor-stage data were missing for too many of these combinations and so I ultimately decided this was probably OK as-is.

```{r showing-missing-tumour-stage}
d %>% dplyr::distinct(specific_treatment, tissue_type, tumour_stage)
```


(code used to check for addition of tumour-stage)
```{r checking-tumour-stage-grade}
mph2 <- update(mph1, . ~ . + tumour_grade)

cox.zph(mph2, transform = log)
mph2.dfbeta <- residuals(mph2, type="dfbeta")
par(mfrow=c(1, 2))
for (j in 1:2) {
  plot(mph2.dfbeta[, j], ylab=names(coef(mph2))[j])
  abline(h=0, lty=2)
  }
title('testing for influential observations', outer = T)
```

(code used to check for addition of tumour-stage)

```{r checking-tumour-stage-cellularity}
mph3 <- update(mph1, . ~ . + percentage_cellularity)
summary(mph3)

# non-prop hazards
cox.zph(mph3, transform = log)

# overly influential?
mph3.dfbeta <- residuals(mph3, type="dfbeta")
par(mfrow=c(1, 2))
for (j in 1:2) {
  plot(mph3.dfbeta[, j], ylab=names(coef(mph3))[j])
  abline(h=0, lty=2)
}
title('testing for influential observations', outer = T)
dev.off() ## reset plot area

# non-linearity?
mph3.martin <- residuals(mph3, type="martingale")
mph3.X <- as.matrix(md %>% tidyr::drop_na(percentage_cellularity) %>% dplyr::select(percentage_cellularity))
plot(mph3.X[, 1], mph3.martin, xlab="percentage_cellularity", ylab="residuals")
title('testing for non-linear assocations', outer = T)
abline(h=0, lty=2)
lines(lowess(mph3.X[, 1], mph3.martin, iter=0))
```

(the above suggests that using bins of cellularity might provide a better measure)

```{r checking-tumour-stage-lcellularity}
mph3a <- update(mph1, . ~ . + level_of_cellularity)
summary(mph3a)

# non-prop hazards
cox.zph(mph3a, transform = log)

# overly influential?
mph3a.dfbeta <- residuals(mph3a, type="dfbeta")
par(mfrow=c(1, 3))
for (j in 1:3) {
  plot(mph3a.dfbeta[, j], ylab=names(coef(mph3a))[j])
  abline(h=0, lty=2)
}
title('testing for influential observations', outer = T)
dev.off() ## reset plot area

# non-linearity?
mph3a.martin <- residuals(mph3a, type="martingale")
mph3a.X <- as.matrix(md %>% tidyr::drop_na(percentage_cellularity) %>% dplyr::select(percentage_cellularity))
plot(mph3a.X[, 1], mph3a.martin, xlab="percentage_cellularity", ylab="residuals")
title('testing for non-linear assocations', outer = T)
abline(h=0, lty=2)
lines(lowess(mph3a.X[, 1], mph3a.martin, iter=0))
```


## Analysis of mutations with survival 

```{r analysis-of-mutations}
mph4 <- update(mph0, . ~ . + tumour_stage + percentage_cellularity + mutations)
summary(mph4)
```

test for non-prop hazards
```{r zph-mutations}
cox.zph(mph4)
cox.zph(mph4, transform = log)
```

test for overly influential observations
```{r infl-mutations}
mph4.dfbeta <- residuals(mph4, type="dfbeta")
par(mfrow=c(1, 3))
for (j in 1:3) {
  plot(mph4.dfbeta[, j], ylab=names(coef(mph4))[j])
  abline(h=0, lty=2)
}
title('test for influential observations', outer = T)
```

Seems likely the above is the observation with v high mutation count.

```{r nonlin-mutations}
mph4.martin <- residuals(mph4, type='martingale')
mph4.X <- as.matrix(md %>% tidyr::drop_na(percentage_cellularity) %>% dplyr::select(percentage_cellularity, mutations))
par(mfrow=c(1, 2))
for (j in 1:2) {
  plot(mph4.X[, j], mph4.martin, xlab=c('percentage_cellularity', 'mutations')[j], ylab="residuals")
  abline(h=0, lty=2)
  lines(lowess(mph4.X[, j], mph4.martin, iter=0))
}
title('test for non-linearity of effect', outer = T)
# this confirms that we should probably drop this observation with very high mutation count
```

Plot KM estimates by mutation count <> median 

```{r km1-mutations}
# plot by mutation count <> median
update(m0, . ~ . + mutations_above_median, data = md) %>%
  survminer::ggsurvplot(., risk.table = TRUE, conf.int = TRUE, pval = TRUE)
```

Plot KM estimate by mutation count <> p75

```{r km2-mutations}
update(m0, . ~ . + mutations_above_p75, data = md) %>%
  survminer::ggsurvplot(., risk.table = TRUE, conf.int = TRUE, pval = TRUE)
```

## Analysis of mutations excluding v high counts

Cox PH analysis excluding outlier mutation count

```{r coxph-mut-excl}
mph4e <- update(mph0, . ~ . + tumour_stage + percentage_cellularity + mutations_excl)
summary(mph4e)
```

test for non-proportional hazards

```{r zph-mut-excl}
# non-prop hazards
cox.zph(mph4e)
cox.zph(mph4e, transform = log)
```

The above suggests we should now stratify by tumour-stage

### Update our analysis to use the stratified analysis

```{r coxph-mut-e-s}
summary(mph4s <- update(mph0, . ~ . + strata(tumour_stage) + percentage_cellularity + mutations_excl))
```

Test for non-prop hazards

```{r zph-mut-e-s}
# non-prop hazards
cox.zph(mph4s)
cox.zph(mph4s, transform = log)
```

Influential observations

```{r inf-mut-e-s}
mph4s.dfbeta <- residuals(mph4s, type="dfbeta")
par(mfrow=c(1, 2))
for (j in 1:2) {
  plot(mph4s.dfbeta[, j], ylab=names(coef(mph4s))[j])
  abline(h=0, lty=2)
}
title('test for influenctial observations', outer = T)
```

Non-linearity of effects 

```{r nonlin-mut-e-s}
mph4s.martin <- residuals(mph4s, type='martingale')
mph4s.X <- as.matrix(md %>% tidyr::drop_na(percentage_cellularity, mutations_excl) %>% dplyr::select(percentage_cellularity, mutations_excl))
par(mfrow=c(1, 2))
for (j in 1:2) {
  plot(mph4s.X[, j], mph4s.martin, xlab=c('percentage_cellularity', 'mutations_excl')[j], ylab="residuals")
  abline(h=0, lty=2)
  lines(lowess(mph4s.X[, j], mph4s.martin, iter=0))
}
title('Test for nonlinear effect', outer = T)
# this confirms that we should probably drop this observation with very high mutation count
```

### How does the model look if we use the binned cellularity data?

```{r analysis-mut-e-s-binnes}
summary(mph5 <- update(mph0, . ~ . + strata(tumour_stage) + level_of_cellularity + mutations_excl))
```

```{r zph-mut-e-s-b}
# non-prop hazards
cox.zph(mph5, transform = log)
```

```{r infl-mut-e-s-b}
# overly influential?
mph5.dfbeta <- residuals(mph5, type="dfbeta")
par(mfrow=c(1, length(coef(mph5))))
for (j in 1:length(coef(mph5))) {
  plot(mph5.dfbeta[, j], ylab=names(coef(mph5))[j])
  abline(h=0, lty=2)
}
```

non-linearity?

```{r nonlin-mut-e-s-b}
mph5.martin <- residuals(mph5, type='martingale')
mph5.X <- as.matrix(md %>% tidyr::drop_na(percentage_cellularity, mutations_excl) %>% dplyr::select(percentage_cellularity, mutations_excl))
par(mfrow=c(1, 2))
for (j in 1:2) {
  plot(mph5.X[, j], mph5.martin, xlab=c('percentage_cellularity', 'mutations_excl')[j], ylab="residuals")
  abline(h=0, lty=2)
  lines(lowess(mph5.X[, j], mph5.martin, iter=0))
}
# this confirms that we should probably drop this observation with very high mutation count
```

### How does the model look if we use the log-cellularity data?

```{r analysis-mut-e-s-log}
summary(mph5b <- update(mph0, . ~ . + strata(tumour_stage) + log1p(percentage_cellularity) + mutations_excl))
```

```{r zph-mut-e-s-l}
# non-prop hazards
cox.zph(mph5b, transform = log)
```

```{r infl-mut-e-s-l}
# overly influential?
mph5b.dfbeta <- residuals(mph5b, type="dfbeta")
par(mfrow=c(1, length(coef(mph5b))))
for (j in 1:length(coef(mph5b))) {
  plot(mph5b.dfbeta[, j], ylab=names(coef(mph5b))[j])
  abline(h=0, lty=2)
}
```

non-linearity?

```{r nonlin-mut-e-s-l}
mph5b.martin <- residuals(mph5b, type='martingale')
mph5b.X <- as.matrix(md %>% tidyr::drop_na(percentage_cellularity, mutations_excl) %>% dplyr::select(percentage_cellularity, mutations_excl))
par(mfrow=c(1, 2))
for (j in 1:2) {
  plot(mph5b.X[, j], mph5b.martin, xlab=c('percentage_cellularity', 'mutations_excl')[j], ylab="residuals")
  abline(h=0, lty=2)
  lines(lowess(mph5b.X[, j], mph5b.martin, iter=0))
}
```

## Analysis of neoantigens (peptides)

```{r coxph-neoant}
summary(mph6 <- update(mph0, . ~ . + strata(tumour_stage) + level_of_cellularity + peptides))
```


non-prop hazards?

```{r zph-neoant}
# non-prop hazards
cox.zph(mph6, transform = log)
```

influence? 

```{r inf-neoant}
mph6.dfbeta <- residuals(mph6, type="dfbeta")
par(mfrow=c(1, length(coef(mph6))))
for (j in 1:length(coef(mph6))) {
  plot(mph6.dfbeta[, j], ylab=names(coef(mph6))[j])
  abline(h=0, lty=2)
}
```

non-linearity?

```{r nonlin-neoant}
mph6.martin <- residuals(mph6, type='martingale')
mph6.X <- as.matrix(md %>% tidyr::drop_na(percentage_cellularity, peptides) %>% dplyr::select(percentage_cellularity, peptides))
par(mfrow=c(1, 2))
for (j in 1:2) {
  plot(mph6.X[, j], mph6.martin, xlab=c('percentage_cellularity', 'peptides')[j], ylab="residuals")
  abline(h=0, lty=2)
  lines(lowess(mph6.X[, j], mph6.martin, iter=0))
}
```

This confirms that we should probably drop this observation with very high mutation count.

## Can we model neoantigens & mutation count effects?

```{r cox-both}
summary(mph7 <- update(mph0, . ~ . + level_of_cellularity + strata(tumour_stage) + mutations_excl + peptides))
```

```{r}
summary(mph7 <- update(mph0, ~ . + level_of_cellularity + strat(tumour_stage) + mutations_excl + peptides))
summary(mph7v2 <- update(mph0, ~ . + level_of_cellularity + strat(tumour_stage) + I(mutations_excl + peptides)))
anova(mph7, mph7v2)
```


